Evaluation of sample enhancment: targeted capture with nanoparticles compared with an automated extraction method

### 2021-12-13 products/manuscript/Manuscript.Rmd
Wordings
Update descriptions in manuscript

---
title: "MADA Data Analysis Project"
subtitle: ""
author: Monica Chan
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::html_document2: 
    toc: false
  bookdown::word_document2: 
    toc: false
    number_sections: true
bibliography: ../references.bib
csl: ../PLoS.csl
---

# Summary/Abstract

The global Coronavirus Disease 2019 (COVID-19) pandemic continues to persist due to the timing of an available vaccine, varying adherence to COVID-19 public health safety recommendations, and the emergence of SARS-CoV-2 (SC2) variants. Environmental reservoirs are a reported, indirect transmission route for viral pathogens. Although many viruses have a low infectious dose, they can be difficult to detect due to the low-bioburden in the environment. The ability to detect viral pathogens is critical to identifying hidden reservoirs which can inform infection control. Culture-based methods can be labor and time intensive. The use and limitations of commercial “target capture” (TC) technologies are varied compared to molecular approaches utilizing automated extraction. The objective of this study is improved outbreak response through rapid pathogen detection by evaluating a TC technology’s ability to capture SC2 from complex matrices.

@@

The global Coronavirus Disease 2019 (COVID-19) pandemic continues to persist due to the timing of an available vaccine, varying adherence to COVID-19 public health safety recommendations, and the emergence of SARS-CoV-2 (SC2) variants. Environmental reservoirs are a reported, indirect transmission route for viral pathogens [@Tanner2021;@Marcenac2021-sz]. Although many viruses have a low infectious dose, they can be difficult to detect due to the low bio-burden in the environment. The ability to detect viral pathogens is critical to identifying hidden reservoirs which can inform infection control [@Paton2021-jc]. Culture-based methods can be labor and time intensive. The use and limitations of commercial "target capture" (TC) technologies are varied compared to molecular approaches utilizing automated extraction. The objective of this study is improve outbreak response through rapid pathogen detection by evaluating a TC technology's ability to capture SC2 from complex matrices.

The virus is capable of human-to-human transmission via inhalation of infected aerosols. Previous evaluations of viral fragment longevity have tested negative and indicate the possibility of no present virus or below assay limit of detection [@Parker2020].

@@

Using established standard methods for environmental sampling with a self-contained polyurethane swab from stainless steel surface, a TC approach (NTP; Nanotrap® Magnetic Virus Particles, Ceres Nanosciences) was compared to an automated extraction method (MX48; Maxwell® RSC Viral Total Nucleic Acid Purification Kit, Promega) for recovery of heat-inactivated SC2 from the polyurethane swab as detected by rRT-PCR. [@Wittwer2012-vt;@Shah2021-mb;@Barclay2020].
Real time RT-PCR was conducted per CDC 2019-nCoV Real-Time RT-PCR Diagnostic Panel protocol to detect the presence of N1 and N2 sequences designed for the specific detection of 2019-nCoV.Reports were QCed and exported from machines electronically.

@@

Data was cleaned and reformatted to relevant factors and numeric data types.
During cleaning the data was split and saved in three forms- All (compilation of all recovery studies- P1 and P2);P1 (only data from spiked studies); and P2 (only data from surface sampling studies).
P1 and P2 were analyzed together due to constant serial dilution controls.
P2 data were expanded on due to the interests of the study focusing on surface recovery.

@@

For Aim 2, data was limited to P2 data from this point onwards to focus on surface recovery of genomic material and limit variability from P1 spiked studies. An exploration in how Tool and Methods differed. Additional exploration occured in looking specifically at the NTP Method and its results as there were additional variables in input.

@@

Serial dilution Ct values increase in a linear manner and within expected intervals indicating log10 loss as the dilution value decreases. Overall the Methods appear to not be too different from one another at higher dilution levels, but we also see that as the Dilution level decreases the the NTP method registers a lower Ct value than the CDC method and begins to differentiate.

@@

Repeat comparisons of the first Aim but with the data from P2 to see if the differences are evident. Overall the Methods appear to not be explicitly different from one another. The divergence of CDC and NTP methods is still seen slightly with CDC recovering less at higher dilution levels and switching places with NTP as the dilution increases.

@@

The larger the input volume the recovered Ct values tends to be smaller. 
Tool A has an average Ct recovery higher than B. This makes sense as the surfaces A recovered from is 2 logs higher (-1) than that of B (-3). A log drop is ~3Ct values we can see from the graphic that it trends as predicted. 
Between methods at an input level of 300ul, there's a lot more variability in the NTP method than the CDC method,.
Probes see little to no difference in Ct Values recovered.

### Phase 2 Comparing Methods with standard input of 300uL

@@

Serial dilution data between CDC and NTP. Noting that all SD are in the confirmed realm of under 27 Ct Value CDC method is better than NTP at -1 concentration of the SD. the -2 shows little difference between method recovery Ct values. NTP appears to remain better than CDC at -3 and -4 levels. Accounting for the SE looks like both methods are the same until the -2 dilution. Might be significant difference at the -4 dilution.

### Phase 2 Nanoparticles only - All tools at 300 and 1000ul input levels

@@

A closer look at specially NTP only results with the CDC input vol and the max vol the manufacturer recommends. The increased input volume appears to result in smaller recovery values than the smaller input volume. Still no difference between Probe types.

@@

Dilution and input play an important part in Ct recovery (\@ref(fig:ImportanceFactor) and \@ref(fig:ImportanceFactor2)). 
Tools A and B are different dilution levels (1:10 and 1:1000, respectively) and the drop between Ct values is consistent with expected log loss equivalent values (\@ref(fig:P2-Tool-AB)). In \@ref(fig:P2-Tool-AB) we can see that as the input increases the general sample has a smaller recovery Ct value. Yet, with the data gathered only after the input is greater than 6800uL does the value surpass the recovery range obtained at the 1000ul input volume. This suggests that the NTP method is able to utilize and capture at least what is available in the sample, and perhaps more.

@@

Data and code required to complete the analysis are available from `MONICACHAN-MADA-project` Repository.

>>>>

---
title: "Evaluation of sample enhancment: targeted capture with nanoparticles compared with an automated extraction method"
subtitle: ""
author: Monica Chan
date: "`r Sys.Date()`"
site: bookdown::bookdown_site
documentclass: article
output:
  bookdown::html_document2: 
    toc: false
  bookdown::word_document2: 
    toc: false
    number_sections: true
bibliography: ../references.bib
csl: ../PLoS.csl
---
# Summary/Abstract

The global Coronavirus Disease 2019 (COVID-19) pandemic continues to persist due to the timing of an available vaccine, varying adherence to COVID-19 public health safety recommendations, and the emergence of SARS-CoV-2 (SC2) variants. Environmental reservoirs are a reported, indirect transmission route for viral pathogens. Although many viruses have a low infectious dose, they can be difficult to detect due to the low-bioburden in the environment. The ability to detect viral pathogens is critical to identifying hidden reservoirs which can inform infection control. Culture-based methods can be labor and time intensive. The use and limitations of commercial “target capture” (TC) technologies are varied compared to molecular approaches utilizing automated extraction. The objective of this study is improved outbreak response through rapid pathogen detection by evaluating a TC technology’s ability to capture SC2 from complex matrices. That data gathered supports that the using TC technology, nanoparticles, does extract from a sample, yet shows no difference between the currently used automated extraction method. Comparisons between the details of method used were also explored. Sample factors such as input and dilution were found to impact sample recovery and environmental data showed very light effects-- as shown by using machine learning methods.

<br><br>

@@

The global Coronavirus Disease 2019 (COVID-19) pandemic continues to persist due to the timing of an available vaccine, varying adherence to COVID-19 public health safety recommendations, and the emergence of SARS-CoV-2 (SC2) variants has encouraged a continued need for better and faster testing.

Culture-based methods can be labor and time intensive. An accepted standard for testing humans for COVID-19 is the SC2 real-time RT-PCR reverse transcription–polymerase chain reaction (RT-PCR)[@Hanson2020]. CT quantification has been shown to have potential for assisting in decision-making in the management of COVID-19 patients[@Hanson2020].

As the virus is capable of human-to-human transmission via inhalation of infected aerosols, the existence of environmental reservoirs is likely. Environmental reservoirs are a known indirect transmission route for viral pathogens [@Tanner2021;@Marcenac2021-sz].The ability to detect it in an environment is critical to identifying hidden reservoirs which can inform infection control [@Paton2021-jc]. Many viruses have a low infectious dose, they can be difficult to detect due to the low bio-burden in the environment.

Previous evaluations of viral fragment longevity have tested negative and indicate the possibility of no present virus or below assay limit of detection [@Parker2020]. The use and limitations of commercial "target capture" (TC) technologies are varied compared to molecular approaches utilizing automated extraction. The objective of this study is improve outbreak response through rapid pathogen detection by evaluating a TC technology's ability to capture SC2 from complex matrices. The TC method proposed to be tested is one based on magnetized hydro-gel particles treated with affinity dyes that can be customized to select for and bind with a multitude of analytes. A retail kit of these particles, NanoTrap Particles, are commercially avalible and advertised as being specially capable of "capturing" the SC2 virus from various sample types such as clinical isolates to waste water better than processing a sample without the use of the particle.

Using RT-PCR we will establish that the TC method, NanoTrap Particles, perform. Then the results will be compared with a standardized automated preparation method. By measuring the Ct values of the  resulting methods we can determine the sensitivity of these extraction methods to capture addition or retain the genomic copies from a sample. The data set will then be broken down and analyzed specifically with surface sampling capture to see if it is possible to recover from the environment with the use of a specialized sampling tool for molecular work.

CT quantification combined with patient symptom data has shown promise when implemented with machine learning, although many studies are focused on patient information or are proof of concept [@Cai2020;@Langer2020;@Gangloff2021]. Additional exploration of the data with some machine learning techniques for any significant importance to Ct value recovery will be performed.
<br>

@@

Using established standard methods for environmental sampling with a self-contained polyurethane swab from stainless steel surface, a TC approach (NTP; NanoTrap® Magnetic Virus Particles, Ceres Nanosciences) was compared to an automated extraction method (MX48; Maxwell® RSC Viral Total Nucleic Acid Purification Kit, Promega) for recovery of heat-inactivated SC2 from the polyurethane swab as detected by RT-PCR. [@Wittwer2012-vt;@Shah2021-mb;@Barclay2020].
Real time RT-PCR was conducted per CDC 2019-nCoV Real-Time RT-PCR Diagnostic Panel protocol to detect the presence of N1 and N2 sequences designed for the specific detection of 2019-nCoV.Reports were QCed and exported from machines electronically.

@@

Data was checked for missing data points (aside from Ct Values) and reformatted factors were mutated to the correct factor or numeric data types.
During cleaning the data was split and saved in three forms- All (compilation of all recovery studies- P1 and P2);P1 (only data from spiked studies); and P2 (only data from surface sampling studies).

P1 and P2 were analyzed together due to constant serial dilution controls.
P2 data were expanded on due to the interests of the study focusing on surface recovery.

Cleaning scripts are found in the `ProcessingScript.Rmd` file in the code folder.

@@

For Aim 2, data was limited to P2 data from this point onward to focus on surface recovery of genomic material and limit variability from P1 spiked studies. An exploration in how Tool and Methods differed. Additional exploration occurred in looking specifically at the NTP Method and its results as there were additional variables in input.

@@

Serial dilution Ct values increase in a linear manner and within expected intervals indicating log10 loss as the dilution value decreases. Overall the Methods appear to not be too different from one another at higher dilution levels, but we also see that as the Dilution level decreases the the NTP method registers a lower Ct value than the CDC method and begins to differentiate. _Dash lines represent thresholds of significance: Orange is the threshold between positive and suspected; Blue line indicates threshold between suspected and detected_.

@@

Repeat comparisons of the first Aim but with the data from P2 to see if the differences are evident. Overall the Methods appear to not be explicitly different from one another. The divergence of CDC and NTP methods is still seen slightly with CDC recovering less at higher dilution levels and switching places with NTP as the dilution increases. _Dash lines represent thresholds of significance: Orange is the threshold between positive and suspected; Blue line indicates threshold between suspected and detected_

@@

The larger the input volume the recovered Ct values tends to be smaller. 
Tool A has an average Ct recovery higher than B. This makes sense as the surfaces A recovered from is 2 logs higher (-1) than that of B (-3). A log drop is ~3Ct values we can see from the graphic that it trends as predicted. 
Between methods at an input level of 300µl, there's a lot more variability in the NTP method than the CDC method,.
Probes see little to no difference in Ct Values recovered. _Dash lines represent thresholds of significance: Orange is the threshold between positive and suspected; Blue line indicates threshold between suspected and detected_

### Phase 2 Comparing Methods with standard input of 300µL

@@

Serial dilution data between CDC and NTP. Noting that all SD are in the confirmed realm of under 27 Ct Value CDC method is better than NTP at -1 concentration of the SD. the -2 shows little difference between method recovery Ct values. NTP appears to remain better than CDC at -3 and -4 levels. Accounting for the SE looks like both methods are the same until the -2 dilution. Might be significant difference at the -4 dilution. _Dash lines represent thresholds of significance: Orange is the threshold between positive and suspected; Blue line indicates threshold between suspected and detected_

### Phase 2 Nanoparticles only - All tools at 300 and 1000µl input levels

@@

A closer look at specially NTP only results with the CDC input vol and the max vol the manufacturer recommends. The increased input volume appears to result in smaller recovery values than the smaller input volume. Still no difference between Probe types. _Dash lines represent thresholds of significance: Orange is the threshold between positive and suspected; Blue line indicates threshold between suspected and detected_

@@

Dilution and input play an important part in Ct recovery (\@ref(fig:ImportanceFactor) and \@ref(fig:ImportanceFactor2)). 
Tools A and B are different dilution levels (1:10 and 1:1000, respectively) and the drop between Ct values is consistent with expected log loss equivalent values (\@ref(fig:P2-Tool-AB)). In \@ref(fig:P2-Tool-AB) we can see that as the input increases the general sample has a smaller recovery Ct value. Yet, with the data gathered only after the input is greater than 6800µL does the value surpass the recovery range obtained at the 1000µl input volume. This suggests that the NTP method is able to utilize and capture at least what is available in the sample, and perhaps more.

@@

Data and code required to complete the analysis are available from `MONICACHAN-MADA-project` [Repository](https://github.com/MoChan2021/MONICACHAN-MADA-project) .
